---
title: "Replication Document"
output: pdf_document
---

This document can be used to replicate the quantitative lithic analyses presented in the paper "Multiple hominin dispersals into Southwest Asia over the last 400,000 years" by Groucutt et al. [2020](http://DOI). All analyses were conducted in R.

#Load libraries
First, we will load the [psych](https://cran.r-project.org/web/packages/psych/psych.pdf) library for convenient PCA-related tests, and the [ggplot2](https://ggplot2.tidyverse.org/) and [ggpubr](https://github.com/kassambara/ggpubr) libraries for plotting results.

```{r}
library(psych)
library(ggplot2)
library(ggpubr)
```

#Data
Next, we load the lithic data as follows:

```{r}
LP <- read.csv(file="./LP.csv")
MIS67 <- read.csv(file="./MIS67.csv")
```

These data sets come with several variables (columns). The Lower Palaeolithic data look like this:

```{r}
head(LP)
```

The data from the transition between Marine Isotope Stage 6 and 7 (MIS67) look lke this:

```{r}
head(MIS67)
```

#PrePCA tests
Before running the analysis, we used a couple of simple preliminary tests to determine whether the variation in the data was sufficiently greater in at least one or more dimensions that sensible principle components could be extracted. One test involved the "Kaiser, Meyer, Olkin Measure of Sampling Adequacy":

```{r}
KMO(MIS67[,c(3:10)])

KMO(LP[,c(3:10)])
```

The other involved "Bartlett's Test for Sphericity",

```{r}
cortest.bartlett(MIS67[,c(3:10)])

cortest.bartlett(LP[,c(3:10)])
```

#PCA
Then, we can perform the simple PCA on the relevant lithic variables,

```{r}
pca_MIS67 <- prcomp(
                x = MIS67[,c(3:10)],
                retx = T,
                center = T,
                scale = T)

pca_LP <- prcomp(
            x = LP[,c(3:10)],
            retx = T,
            center = T,
            scale = T)

LP_scores <- cbind(
                LP[,c(1:2)],
                pca_LP$x)

MIS67_scores <- cbind(
                    MIS67[,c(1:2)],
                    pca_MIS67$x)
```

#Ploting
Lastly, we plot the results using ggplot2 as follows. The first plot will contain the results for the analysis of the LP data,

```{r label="PCA Results for the LP data"}
sample_name <- "LP"

p1 <- ggplot(
        data = get(paste(sample_name,"_scores",sep="")),
        mapping = aes(Assemblage,PC1,group = Assemblage)) +
      geom_boxplot(colour="darkgrey",fill="grey",alpha=0.8) +
      theme_minimal() +
      theme(text = element_text(family="Times", size=12),
            plot.title = element_text(face="bold",hjust=0.5,size=15),
            axis.text.x = element_blank(),
            axis.title.x = element_blank())

p2 <- ggplot(
        data = get(paste(sample_name,"_scores",sep="")),
        mapping = aes(Assemblage,PC2,group = Assemblage)) +
      geom_boxplot(colour="darkgrey",fill="grey",alpha=0.8) +
      theme_minimal() +
      theme(text = element_text(family="Times", size=12),
            plot.title = element_text(face="bold",hjust=0.5,size=15),
            axis.text.x = element_blank(),
            axis.title.x = element_blank())

p3 <- ggplot(
        data = get(paste(sample_name,"_scores",sep="")),
        mapping = aes(Assemblage,PC3,group = Assemblage)) +
      geom_boxplot(colour="darkgrey",fill="grey",alpha=0.8) +
      theme_minimal() +
      theme(text = element_text(family="Times", size=12),
            plot.title = element_text(face="bold",hjust=0.5,size=15),
            axis.text.x = element_text(size=8),
            axis.title.x = element_blank())

fig <- ggarrange(p1,p2,p3,
         ncol=1,
         nrow=3,
         align="v")

annotate_figure(fig,
               top = text_grob("PCA Score Box Plots\nLP",
                                 family="Times",
                                 face="bold"),
               fig.lab.pos = "top")
```

The second plot contains the results pertaining to the MIS67 data,

```{r label="PCA Results for the MIS67 data"}
sample_name <- "MIS67"

p1 <- ggplot(
        data = get(paste(sample_name,"_scores",sep="")),
        mapping = aes(Assemblage,PC1,group = Assemblage)) +
      geom_boxplot(colour="darkgrey",fill="grey",alpha=0.8) +
      theme_minimal() +
      theme(text = element_text(family="Times", size=12),
            plot.title = element_text(face="bold",hjust=0.5,size=15),
            axis.text.x = element_blank(),
            axis.title.x = element_blank())

p2 <- ggplot(
        data = get(paste(sample_name,"_scores",sep="")),
        mapping = aes(Assemblage,PC2,group = Assemblage)) +
      geom_boxplot(colour="darkgrey",fill="grey",alpha=0.8) +
      theme_minimal() +
      theme(text = element_text(family="Times", size=12),
            plot.title = element_text(face="bold",hjust=0.5,size=15),
            axis.text.x = element_blank(),
            axis.title.x = element_blank())

p3 <- ggplot(
        data = get(paste(sample_name,"_scores",sep="")),
        mapping = aes(Assemblage,PC3,group = Assemblage)) +
      geom_boxplot(colour="darkgrey",fill="grey",alpha=0.8) +
      theme_minimal() +
      theme(text = element_text(family="Times", size=12),
            plot.title = element_text(face="bold",hjust=0.5,size=15),
            axis.text.x = element_text(size=8),
            axis.title.x = element_blank())

fig <- ggarrange(p1,p2,p3,
         ncol=1,
         nrow=3,
         align="v")

annotate_figure(fig,
               top = text_grob("PCA Score Box Plots\nMIS67",
                                 family="Times",
                                 face="bold"),
               fig.lab.pos = "top")
```
